Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-09-02 main.Rmd
Section 6

```{r linear-time, echo=FALSE, out.width = "100%", fig.cap="(ref:linear-time)", fig.pos="!htb"}
```

@@

```{r circular-dow,echo=FALSE,out.width = "100%",fig.cap= "(ref:circular-dow)",fig.pos="!htb"}
```

@@

```{r aperiodic-example, echo=FALSE, out.width = "100%", fig.cap= "(ref:aperiodic-example)", fig.pos="!htb"}
```

@@

```{r allFig, out.width="100%", fig.pos="!htb", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap= "(ref:allFig)", fig.show = 'hold'}
```

@@

Smart meters provide large quantities of measurements on energy usage for households across Australia. One of the customer trials [@smart-meter] conducted as part of the Smart Grid Smart City project in Newcastle, New South Wales and some parts of Sydney provides customer wise data on energy consumption for every half hour from February 2012 to March 2014. The idea here is to show how to visualize the distribution of the energy consumption across different cyclic granularities in a systematic way to identify different behavioral patterns.

@@

The tsibble object `smart_meter10` from R package `gravitas` [@R-gravitas] consisting of `reading_datetime`, `customer_id` and `general_supply_kwh` denoting the index, key and measured variable of the tsibble is used to facilitate the systematic exploration. While trying to explore the energy behavior of these customers systematically across cyclic time granularities, the first thing to consider is which cyclic time granularities we can look at exhaustively. Let us consider conventional time deconstructions for a Gregorian calendar (second, minute, half-hour, hour, day, week, month, year). Since the interval of this tsibble is 30 minutes, the temporal granularities may range from half-hour to year. Considering $6$ linear granularities half-hour, hour, day, week, month and year in the hierarchy table, $N_C = (6*5/2) = 15$. If $N_C$ seem too large, the smallest and largest linear granularities could be considered to be removed for analysis. Granularities half-year and year are removed to have $N_C = (4*3/2) = 6$ and obtain cyclic granularities namely "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month", read as "hour of the day", etc. Further, we add cyclic granularity day-type( "wknd_wday") to capture weekend and weekday behavior. Now that we have a list of cyclic granularities to look at, we should be able to compute them using Section&nbsp;\ref{sec:cyclic-calendar}.

```{r search, echo=FALSE, eval = FALSE}
```

```{r search_gran_limit2, echo = FALSE, eval = FALSE}
```

@@

From the search list, $N_C = 7$ cyclic granularities are chosen for which we would like to derive insights of energy behavior. Recalling the data structure <$C_i$, $C_j$, `general_supply_kwh`> for exploration $\forall i, j \in \{1, 2, \ldots, 7\}$, each of these $7$ cyclic granularities can either be mapped to x-axis or to facet. Choosing $2$ of the possible $7$ granularities, which is equivalent to having $^{7}P_2 = 42$ candidates for visualization. Harmonies can be identified among those $42$ possibilities to narrow the search. \autoref{tab:harmony-tab} shows $16$ harmony pairs after removing clashes and any cyclic granularities with levels more than $31$, as effective exploration becomes difficult with many levels (Section&nbsp;\ref{sec:aesthetics}).

@@

```{r bothcust, fig.cap = "(ref:bothcust)", out.width="90%", fig.pos="!htb"}
```

(ref:bothcust) Energy consumption of a single customer shown with different distribution displays, and granularity arrangements. Two granularities are used: hour of the day (I) and weekday/weekend (II). Plot (a) shows granularity I faceted by granularity II, and plots (b), (c) shows the converse mapping. Plot (a) makes a comparison of usage by workday within each hour of the day using side-by-side boxplots. Generally, on a work day there is more consumption early in the day. Plots (b) and (c) examine the temporal trend of consumption over the course of a day, separately for the type of day. Plot (b) uses an area quantile to put the emphasis on the time series, for example, the median consumption over time shows prolonged usage in the morning on weekdays. Plot (c) uses a violin plot to place emphasis on distributional differences across hours. It can be seen that the morning use on weekdays is bimodal, some work days there is low usage, which might indicate the person is working from home and also having a late start.

Few harmony pairs are displayed in \autoref{fig:bothcust} to illustrate the impact of different distribution plots and reverse mapping. For each of \autoref{fig:bothcust} (b) and (c), $C_i$ is the circular granularity day-type (weekday/weekend) and $C_j$ is hour of the day. The geometry used for displaying the distribution is chosen as area-quantiles and violins in \autoref{fig:bothcust} (b and c respectively). \autoref{fig:bothcust} (a) displays reverse mapping of $C_i$ and $C_j$ with $C_i$ denoting hour of the day and $C_j$ denoting day-type with distribution geometrically displayed as boxplots.
In \autoref{fig:bothcust} (b), the black line is the median, whereas the purple band covers 25th to 75th percentile, the orange band covers 10th to 90th percentile and the green band covers 1st to 99th percentile. The first facet represents the weekday behavior while the second one displays the weekend behavior and energy consumption across each hours of the day is shown inside each facet. The energy consumption is extremely (positive- or right-) skewed with the 1st, 10th and 25th percentile lying relatively close whereas 75th, 90th and 99th lying further away from each other. This is common across both weekdays and weekends. For the first few hours on weekdays, median energy consumption starts and continues to be higher for longer as compared to weekends.

Consider looking at violin plots instead of quantile plots to look at the same data in \autoref{fig:bothcust}(c) for some more insights. There is bimodality in the early hours of the day, implying both low and high energy consumption is probable in the early hours of the day both for weekdays and weekends. If we visualize the same data with reverse mapping of the cyclic granularities, then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. For example in \autoref{fig:bothcust}(a), it can be seen that median energy consumption for the early morning hours is higher for weekdays than weekends. Also, outliers are more prominent in the latter hours of the day. All of these indicate that looking at different distribution geometry or changing the mapping might shed lights on different aspect of the energy behavior for the same sample population.

If the data for all keys are visualized together, it might lead to Simpson's paradox, which occurs when one observation shows a particular behavior, but this behavior paradoxically becomes obscured by aggregation. For example in a particular neighborhood one household may have the least daily power consumption for a full week, yet still not be the household with the minimum weekly power consumption. This is an intuitive possibility, because heterogeneous `customer_id`'s with very different occupation or demographics will tend to have very different energy behavior and combining them together will somehow weaken any typical or extreme behavior. A strategy for analyzing periodicities of multiple keys together could be to group the keys basis their energy behavior across different cyclic granularities. This is beyond the scope of the current work.

## T20 cricket data of Indian Premiere League {#sec:cricket}

The method is not only restricted to temporal data, and can be generalized to many hierarchical granularities (with continuous and uni-directional nature). We illustrate this with an application to the sport cricket. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In the Twenty20 format, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus, similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. The idea of cyclic granularities can be likewise mapped to this hierarchy. Example granularities then include ball of the over, over of the innings and ball of the innings. Although most of these cyclic granularities are circular in design of the hierarchy, in application of the rules some granularities are aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an innings finishes before the over finishes. Thus, the cyclic granularity ball-of-over will be circular in most cases and aperiodic in others.

The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. The ball by ball data for IPL season 2008 to 2016 is fetched from [Kaggle](https://www.kaggle.com/josephgpinto/ipl-data-analysis/data). The `cricket` data set in the `gravitas` package summarizes the ball-by-ball data across overs and contains information for a sample of 214 matches spanning 9 seasons (2008 to 2016) such that each over has 6 balls, each innings has 20 overs and each match has 2 innings. This could be useful in a periodic world when we wish to compute any circular/quasi-circular granularity based on a hierarchy table which look like \autoref{tab:hierarchy-cric}. However, even if the situation is not periodic, it can be interesting to visualize the distribution of a measured variable to shed light on the aperiodic behavior of a non-temporal data set similar to aperiodic events like formal meetings, workshops, conferences, school semesters in a temporal set up.

@@

There are many interesting questions that could possibly be answered with the `cricket` data set. For example, it would be interesting to see if the distribution of total runs vary depending on if a team bats in the first or second innings. The Mumbai Indians (MI) and Chennai Super kings (CSK) appeared in final playoffs from 2010 to 2015. We take their example in order to dive deeper into this question. From Figure \ref{fig:cricex}(a), it can be observed that for the team batting in the first innings there is an upward trend of runs per over, while there is no clear upward trend in median and quartile deviation of runs for the teams batting in the second inning. This seem to indicate that players feel mounting pressure to score more runs as they approach towards the end of the first inning. Whereas teams batting in the second innings have a set target in mind and are not subjected to such mounting pressure and may adopt a more conservative strategy, to score runs. Thus winning teams like CSK and MI seem to employ different inning strategies when it comes to their batting order.

Another fascinating exploration would be to examine if runs per over decrease in the subsequent over if fielding (defending) was good in the previous over? For establishing the fielding quality, we apply an indicator function on dismissals (1 if there was at least one wicket in the previous over due to run out or catch, 0 otherwise). Runs in the current over is then the observation variable. Dismissals in the previous over can lead to a batsman adopting a more defensive play style. Figure \ref{fig:cricex}(b) shows that no dismissals in the previous over leads to a higher median and quartile spread of runs per over as compared to the case when there has been at least one dismissal in the previous over. Wickets per over are considered as an aperiodic cyclic granularity with wickets as an aperiodic linear granularity. These granularities do not appear in the hierarchy table since it is difficult to position them in a hierarchy. These are similar to holidays or special events in temporal data.

```{r cricex, fig.cap= "(ref:cricex)", warning = FALSE, message = FALSE, out.width = "90%"}
```

(ref:cricex) Runs per over shown with different distribution displays, and granularities. Plot (a) shows letter value plot across overs faceted by innings. For the team batting in the first innings there is an upward trend of runs per over, while there is no such pattern of runs for the teams batting in the second innings. Plot (b) shows quantile plot of runs per over across an indicator of wickets in previous over faceted by current over. This indicates that at least one wicket in the previous over leads to lower median run rate and quartile spread in the subsequent over.

>>>>

```{r linear-time, echo=FALSE, fig.cap="(ref:linear-time)", fig.pos="!htb"}
```

@@

```{r circular-dow, echo=FALSE, fig.cap= "(ref:circular-dow)",fig.pos="!htb"}
```

@@

```{r aperiodic-example, echo=FALSE, fig.cap= "(ref:aperiodic-example)", fig.pos="!htb"}
```

@@

```{r allFig, fig.pos="!htb", fig.align= 'left',echo=FALSE, eval=TRUE, warning=FALSE,message = FALSE, fig.cap= "(ref:allFig)", fig.show = 'hold'}
```

@@

Smart meters provide large quantities of measurements on energy usage for households across Australia. One of the customer trials [@smart-meter] conducted as part of the Smart Grid Smart City project in Newcastle and parts of Sydney provides customer level data on energy consumption for every half hour from February 2012 to March 2014. We can use this data set to visualize the distribution of energy consumption across different cyclic granularities in a systematic way to identify different behavioral patterns.

@@

The tsibble object `smart_meter10` from R package `gravitas` [@R-gravitas] includes the variables `reading_datetime`, `customer_id` and `general_supply_kwh` denoting the index, key and measured variable respectively. The interval of this tsibble is 30 minutes.

To identify the available cyclic time granularities, consider the conventional time deconstructions for a Gregorian calendar that can be formed from the 30-minute time index: half-hour, hour, day, week, month, quarter, half-year, year. In this example, we will consider the granularities hour, day, week and month giving six cyclic granularities "hour_day", "hour_week", "hour_month", "day_week", "day_month" and "week_month", read as "hour of the day", etc. To these we add day-type ("wknd_wday") to capture weekend and weekday behavior. Now that we have a list of cyclic granularities to look at, we can compute them using the results in Section&nbsp;\ref{sec:cyclic-calendar}.

@@

Using these seven cyclic granularities, we want to explore patterns of energy behavior. Each of these seven cyclic granularities can either be mapped to the x-axis or to facets. Choosing $2$ of the possible $7$ granularities, gives $^{7}P_2 = 42$ candidates for visualization. Harmonies can be identified among those $42$ possibilities to narrow the search. \autoref{tab:harmony-tab} shows $16$ harmony pairs after removing clashes and any cyclic granularities with more than $31$ levels, as effective exploration becomes difficult with many levels (Section&nbsp;\ref{sec:aesthetics}).

@@

```{r bothcust, fig.cap = "(ref:bothcust)", fig.pos="!htb"}
```

(ref:bothcust) Energy consumption of a single customer shown with different distribution displays, and granularity arrangements. Two granularities are used: 1. hour of the day; and 2. weekday/weekend. Plot a shows granularity 1 faceted by granularity 2; plots b and c shows the reverse mapping. Plot a allows comparison of usage by workday within each hour of the day using side-by-side boxplots, showing that on work days there is more consumption early in the day. Plots b and c examine the temporal trend of consumption over the course of a day, separately for the type of day. Plot b uses an area quantile to emphasise the time component; e.g., median consumption shows prolonged high usage in the morning on weekdays. Plot c uses a violin plot emphasise subtler distributional differences across hours, showing that morning use on weekdays is bimodal, and on some work days there is low usage possibly indicating the person is working from home or having a late start.

A few harmony pairs are displayed in \autoref{fig:bothcust} to illustrate the impact of different distribution plots and reverse mapping. For each of \autoref{fig:bothcust}b and c, $C_i$ denotes day-type (weekday/weekend) and $C_j$ is hour-of-day. The geometry used for displaying the distribution is chosen as area-quantiles and violins in \autoref{fig:bothcust}b and c respectively. \autoref{fig:bothcust}a shows the reverse mapping of $C_i$ and $C_j$ with $C_i$ denoting hour-of-day and $C_j$ denoting day-type with distribution geometrically displayed as boxplots.

In \autoref{fig:bothcust}b, the black line is the median, whereas the purple band covers the 25th to 75th percentile, the orange band covers the 10th to 90th percentile, and the green band covers the 1st to 99th percentile. The first facet represents the weekday behavior while the second facet displays the weekend behavior; energy consumption across each hour of the day is shown inside each facet. The energy consumption is extremely skewed with the 1st, 10th and 25th percentile lying relatively close whereas 75th, 90th and 99th lying further away from each other. This is common across both weekdays and weekends. For the first few hours on weekdays, median energy consumption starts and continues to be higher for longer compared to weekends.

The same data is shown using violin plots instead of quantile plots in \autoref{fig:bothcust}c. There is bimodality in the early hours of the day for weekdays and weekends. If we visualize the same data with reverse mapping of the cyclic granularities (\autoref{fig:bothcust}a), then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. Then it can be seen that median energy consumption for the early morning hours is higher for weekdays than weekends. Also, outliers are more prominent in the latter hours of the day. All of these indicate that looking at different distribution geometry or changing the mapping can shed light on different aspects of energy behavior for the same sample.

## T20 cricket data of Indian Premier League {#sec:cricket}

Our proposed approach can be generalized to other hierarchical granularities where there is an underlying ordered index. We illustrate this with data from the sport cricket. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering over the course of the game. In the Twenty20 format, an over will consist of 6 balls (with some exceptions), an innings is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus, there is a hierarchy where ball is nested within overs, overs nested within innings, and innings within matches. Cyclic granularities can be constructed using this hierarchy. Example granularities include ball of the over, over of the innings, and ball of the innings. The hierarchy table is given in  \autoref{tab:hierarchy-cric}.

@@

Although most of these cyclic granularities are circular by design of the hierarchy, in practice some granularities are aperiodic. For example, most overs will consist of 6 balls, but there are exceptions due to wide balls, no-balls, or when an innings finishes before the over finishes. Thus, the cyclic granularity ball-of-over may be aperiodic.

The Indian Premier League (IPL) is a professional Twenty20 cricket league in India contested by eight teams representing eight different cities in India. The IPL ball-by-ball data is provided in the `cricket` data set in the `gravitas` package for a sample of 214 matches spanning 9 seasons (2008 to 2016) such that each over has 6 balls, each innings has 20 overs and each match has 2 innings.

There are many interesting questions that could be addressed with the `cricket` data set. For example, does the distribution of total runs vary depending on if a team bats in the first or second innings? The Mumbai Indians (MI) and Chennai Super Kings (CSK) appeared in final playoffs from 2010 to 2015. Using  data from these two teams, it can be observed (\autoref{fig:cricex}a) that that for the team batting in the first innings there is an upward trend of runs per over, while there is no clear upward trend in median and quartile deviation of runs for the team batting in the second innings after the first few overs. This suggests that players feel mounting pressure to score more runs as they approach the end of the first innings, while teams batting second have a set target in mind and are not subjected to such mounting pressure and therefore may adopt a more conservative run-scoring strategy.

Another question that can be addressed is if good fielding (defending) affects the scoring rate in the subsequent over? To measure the fielding quality, we use an indicator function on dismissals (1 if there was at least one wicket in the previous over due to a run out or catch, 0 otherwise). The scoring rate is measured by runs per over. \autoref{fig:cricex}b shows that no dismissals in the previous over leads to a higher median and quartile spread of runs per over compared to the case when there has been at least one dismissal in the previous over. Thus, dismissals lead a batsman to adopt a more defensive play style. This seems to be unaffected by the over of the innings (the facet variable).

Here, wickets per over is an aperiodic cyclic granularity, so it does not  appear in the hierarchy table. These are similar to holidays or special events in temporal data.

```{r cricex, fig.cap= "(ref:cricex)", warning = FALSE, message = FALSE, fig.height=4, fig.width=8}
```

(ref:cricex) Runs per over shown with different distribution displays, and granularities. Plot a shows letter value plots across overs faceted by innings. For the team batting in the first innings there is an upward trend of runs per over, while there is no such pattern of runs for the teams batting in the second innings. Plot b shows quantile plots of runs per over across an indicator of wickets in the previous over faceted by current over. This indicates that at least one wicket in the previous over leads to lower median run rate and quartile spread in the subsequent over, regardless of the over of the innings.
